Statistical Model Selection for Markov Models of Biomolecular Dynamics

### 2014-04-07 paper/paper2.tex
small changes

\section{Background}
Let $\mathbf{x}_t$ be a ergodic time-homogenous Markov process the phase space $\Omega$ with stationary density $\mu(\mathbf{x})$. Consider an ensemble of such processes at time $t$, described by a distribution $p_t(\mathbf{x})$. There exists an operator, termed the propagator, that evolves the a distribution forward in time at discrete intervals, $\tau$.
\begin{equation}
p_{t+\tau}(\mathbf{x}) = \mathcal{Q}^{(\tau)} \circ p_{t}(\mathbf{x})
\end{equation}
The eigenfunctions of the propagator form a basis for $p(\mathbf{x})$, and the time evolution of the ensemble after $n$ applications of the propagator can be decomposed into a sum of terms along each eigenfunction.
\begin{equation}
\label{eq:prop-timescales}
p_{t+n\tau}(\mathbf{x}) = \sum_{i=1}^\infty \exp\left(-\frac{n \cdot \tau}{t_i}\right) \langle \phi_i, p_t \rangle_{\mu^{-1}} \phi_i(\mathbf{x}),
\end{equation}
where $t_i = -\displaystyle\frac{\tau}{\ln \lambda_i}$ are the propagator's relaxation timescales and $\phi_i$ and $\lambda_i$ are the eigenfunctions and eigenvalues of the propagator, and $\langle \cdot, \cdot \rangle_{\mu^{-1}}$ denotes the inner product defined as
\begin{equation}
\langle \phi_i, p_t \rangle_{\mu^{-1}} = \int_\Omega d\mathbf{x}\; \phi_i(\mathbf{x}) p_t(\mathbf{x}) \frac{1}{\mu(\mathbf{x})}.
\end{equation}
For a more detailed discussion of the propagator and its properties, we refer the reader to \citet{Prinz2011Markov}.

@@

\section{The Bias-Variance Dilemma}
For MSM methods to be valuable to practitioners, they must operate in limited data regimes. Thus, we seek a method for model selection that is sensitive to both the systematic error discussed above, as well as the statistical error associated with estimating the model's parameters. Whereas model selection based on the variational principle described above would tend to increase the number of states without bound, as the magnitude of the eigenvalue bias decreases with the number of states, practical application of this criteria would cause a catastrophic increase in statistical uncertainty in the model. This tradeoff between low-variance (small $k$) but biased estimators, and high-variance (large $k$) but unbiased estimators is a general feature of problems in statistical learning\cite{Sammut2010Encyclopedia}. For example, it can be shown that the uncertainty in the posterior distribution of $\mathbf{T}$ given $S$ and $D$ approaches infinity as $k$ approaches infinity.

@@

\section{Likelihood of a Markov State Model}
Bayes' rule provides a foundation for model selection, by establishing the proportionality of the probability of a model given the data to the probability of the data given the model. Without a strong prior, model selection is reduced to the search for the model that maximizes the likelihood of the data. 
With $\{\mathbf{x}_t\}_{t=1}^N$ an observed trajectory of length $N$, the likelihood can be written as:
\begin{equation}
P\big(\{\mathbf{x}_t\}_{t=1}^N\;\big|\; \{S, T\}\big)d\mathbf{x}^N = \pi_{\sigma(\mathbf{x}_1)}\prod_{t=1}^{N-1} T_{\sigma(\mathbf{x}_t), \sigma(\mathbf{x}_{t+1})} \cdot \prod_{t=1}^{N} P(\mathbf{x}_t | \sigma(\mathbf{x}_t)) \cdot d\mathbf{x}^N,
\label{eq:like}
\end{equation} 
where $\pi_i$ is the stationary probability of state $i$ according to $\mathbf{T}$ and $P(\mathbf{x}_t | \sigma(\mathbf{x}_t))$ is the probability of sampling a conformation $\mathbf{x}_t$ given that the process is in the state $\sigma(\mathbf{x}_t)$, referred to as the emission distribution of state $\sigma(\mathbf{x}_t)$. There is some flexibility in the choice of emission distribution, though \cref{eq:like} is valid only when the emission distributions satisfy
\begin{equation}
P(\mathbf{x} | i) = 0 \;\;\;\; \forall \; \mathbf{x} \textnormal{ s.t. } \sigma(\mathbf{x}) \neq i.
\end{equation}
Given the coarse-graining of $p(\mathbf{x})$ in \cref{eq:coarse-grain-p}, the most natural choice emission distributions are normalized indicator functions
\begin{equation}
P(\mathbf{x} \;|\; i) = \frac{1}{V_i} \mathbf{1}_{s_i}(\mathbf{x}).
\end{equation}
An alternative emission distribution was proposed by \citet{Kellogg2012Evaluation} Instead of having support on $\Omega$, this distribution is supported on the training set $\{\mathbf{x}_t\}_{t=1}^N$. It is given by
\begin{equation}
P(\mathbf{x} \;|\; i) = \big| \{ \mathbf{x}_k : \sigma(\mathbf{x}_k) = \sigma(\mathbf{x}) \} \big|^{-1} \sum_{t=1}^N \delta(\mathbf{x}-\mathbf{x}_t),
\end{equation} where $| \cdot |$ denotes the cardinality of a set and $\delta(\cdot)$ is the Dirac delta function. As this is a discriminative model, it is unable to generalize and assign probability to new data. In certain circumstances, this is an undesirable property. For example, protocols that involve fitting and validating models on separate data sets (e.g. cross-validation) are impossible when the statistical model lacks the capacity to describe new data.

\section{Statistical Model Selection}
Likelihood maximization on the training data (i.e. the data used to fit the model) is insufficient for model selection when the number of parameters varies between proposed models, as more complex models generally exhibit higher empirical likelihoods, often at the cost of larger generalization errors due to over-fitting\cite{Liddle2007Information, Hastie01Elements}. Statistical learning theory provides a number of methods to overcome this problem. Conceptually, the most straightforward is a full Bayesian treatment in which all unknown model parameters are represented by probability distributions. The evidence for a $k$-state model is computed by formally integrating \cref{eq:bayes} over the model parameters and the evidence ratio, or Bayes factor\cite{Gelfand94Bayesian}, then provides a rigorous basis of model selection that appropriately punishes overly complex models as they become poorly constrained in parameter space. Unfortunately such approaches are impractical for problems of this size because of the need to integrate over all possible Markov models of a given size.

@@

@@

\begin{acknowledgement}
It is an honor and a pleasure to contribute this paper to the Swope Festschrift.  I (V.S.P.) have been working with Bill since my early days at Stanford and have benefitted greatly from our numerous interactions with him.  Bill's insight, clarity of thought, and careful approach has influenced my lab's work very broadly.  I am very pleased that our collaborations continue through today and look forward to many more in the future.
\end{acknowledgement}

>>>>

\section{Background}
\subsection{Markov State Models}
Let $\mathbf{x}_t$ be a ergodic time-homogenous Markov process the phase space $\Omega$ with stationary density $\mu(\mathbf{x})$. Consider an ensemble of such processes at time $t$, described by a distribution $p_t(\mathbf{x})$. There exists an operator, termed the propagator, that evolves the a distribution forward in time at discrete intervals, $\tau$.
\begin{equation}
p_{t+\tau}(\mathbf{x}) = \mathcal{Q}^{(\tau)} \circ p_{t}(\mathbf{x})
\end{equation}
The eigenfunctions of the propagator form a basis for $p(\mathbf{x})$, and the time evolution of the ensemble after $n$ applications of the propagator can be decomposed into a sum of terms along each eigenfunction.
\begin{equation}
\label{eq:prop-timescales}
p_{t+n\tau}(\mathbf{x}) = \sum_{i=1}^\infty \exp\left(-\frac{n \cdot \tau}{t_i}\right) \langle \phi_i, p_t \rangle_{\mu^{-1}} \phi_i(\mathbf{x}),
\end{equation}
where $t_i = -\displaystyle\frac{\tau}{\ln \lambda_i}$ are the propagator's relaxation timescales and $\phi_i$ and $\lambda_i$ are the eigenfunctions and eigenvalues of the propagator, and $\langle \cdot, \cdot \rangle_{\mu^{-1}}$ denotes the inner product defined as
\begin{equation}
\langle \phi_i, p_t \rangle_{\mu^{-1}} = \int_\Omega d\mathbf{x}\; \phi_i(\mathbf{x}) p_t(\mathbf{x}) \frac{1}{\mu(\mathbf{x})}.
\end{equation}
For a more detailed discussion of the propagator and its properties, we refer the reader to \citet{Prinz2011Markov}.

@@

\subsection{The Bias-Variance Dilemma}
For MSM methods to be valuable to practitioners, they must operate in limited data regimes. Thus, we seek a method for model selection that is sensitive to both the systematic error discussed above, as well as the statistical error associated with estimating the model's parameters. Whereas model selection based on the variational principle described above would tend to increase the number of states without bound, as the magnitude of the eigenvalue bias decreases with the number of states, practical application of this criteria would cause a catastrophic increase in statistical uncertainty in the model. This tradeoff between low-variance (small $k$) but biased estimators, and high-variance (large $k$) but unbiased estimators is a general feature of problems in statistical learning\cite{Sammut2010Encyclopedia}. For example, it can be shown that the uncertainty in the posterior distribution of $\mathbf{T}$ given $S$ and $D$ approaches infinity as $k$ approaches infinity.

@@

\subsection{Likelihood of a Markov State Model}
Bayes' rule provides a foundation for model selection, by establishing the proportionality of the probability of a model given the data to the probability of the data given the model. Without a strong prior, model selection is reduced to the search for the model that maximizes the likelihood of the data. 
With $\{\mathbf{x}_t\}_{t=1}^N$ an observed trajectory of length $N$, the likelihood can be written as:
\begin{equation}
P\big(\{\mathbf{x}_t\}_{t=1}^N\;\big|\; \{S, T\}\big)d\mathbf{x}^N = \pi_{\sigma(\mathbf{x}_1)}\prod_{t=1}^{N-1} T_{\sigma(\mathbf{x}_t), \sigma(\mathbf{x}_{t+1})} \cdot \prod_{t=1}^{N} P(\mathbf{x}_t | \sigma(\mathbf{x}_t)) \cdot d\mathbf{x}^N,
\label{eq:like}
\end{equation} 
where $\pi_i$ is the stationary probability of state $i$ according to $\mathbf{T}$ and $P(\mathbf{x}_t | \sigma(\mathbf{x}_t))$ is the probability of sampling a conformation $\mathbf{x}_t$ given that the process is in the state $\sigma(\mathbf{x}_t)$, referred to as the emission distribution of state $\sigma(\mathbf{x}_t)$. There is some flexibility in the choice of emission distribution, though \cref{eq:like} is valid only when the emission distributions satisfy
\begin{equation}
P(\mathbf{x} | i) = 0 \;\;\;\; \forall \; \mathbf{x} \textnormal{ s.t. } \sigma(\mathbf{x}) \neq i.
\end{equation}
Given the coarse-graining of $p(\mathbf{x})$ in \cref{eq:coarse-grain-p}, the most natural choice emission distributions are normalized indicator functions
\begin{equation}
P(\mathbf{x} \;|\; i) = \frac{1}{V_i} \mathbf{1}_{s_i}(\mathbf{x}).
\end{equation}
An alternative emission distribution was proposed by \citet{Kellogg2012Evaluation} Instead of having support on $\Omega$, this distribution is supported on the training set $\{\mathbf{x}_t\}_{t=1}^N$. It is given by
\begin{equation}
P(\mathbf{x} \;|\; i) = \big| \{ \mathbf{x}_k : \sigma(\mathbf{x}_k) = \sigma(\mathbf{x}) \} \big|^{-1} \sum_{t=1}^N \delta(\mathbf{x}-\mathbf{x}_t),
\end{equation} where $| \cdot |$ denotes the cardinality of a set and $\delta(\cdot)$ is the Dirac delta function. As this is a discriminative model, it is unable to generalize and assign probability to new data. In certain circumstances, this is an undesirable property. For example, protocols that involve fitting and validating models on separate data sets (e.g. cross-validation) are impossible when the statistical model lacks the capacity to describe new data.

\subsection{Statistical Model Selection}
Likelihood maximization on the training data (i.e. the data used to fit the model) is insufficient for model selection when the number of parameters varies between proposed models, as more complex models generally exhibit higher empirical likelihoods, often at the cost of larger generalization errors due to over-fitting\cite{Liddle2007Information, Hastie01Elements}. Statistical learning theory provides a number of methods to overcome this problem. Conceptually, the most straightforward is a full Bayesian treatment in which all unknown model parameters are represented by probability distributions. The evidence for a $k$-state model is computed by formally integrating \cref{eq:bayes} over the model parameters and the evidence ratio, or Bayes factor\cite{Gelfand94Bayesian}, then provides a rigorous basis of model selection that appropriately punishes overly complex models as they become poorly constrained in parameter space. Unfortunately such approaches are impractical for problems of this size because of the need to integrate over all possible Markov models of a given size.

@@

For all models, we computed the BIC and AIC according to \cref{eq:bic} and \cref{eq:aid}. For the BIC, we need to define the ``number of data points" in the dataset. We used the total number of transitions observed in the data as the value for $N$. However, as transitions were counted using the ``sliding window" approach, these transitions are not necessarily independent. Therefore, the BIC may be over-penalizing the number of parameters due to this over-estimate of $N$. For a discussion of the sliding window approach, we refer the reader to \citet{Prinz2011Markov}.

@@

\begin{acknowledgement}
It is an honor and a pleasure to contribute this paper to the Swope Festschrift.  I (V.S.P.) have been working with Bill since my early days at Stanford and have benefitted greatly from our numerous interactions with him.  Bill's insight, clarity of thought, and careful approach has influenced my lab's work very broadly.  I am very happy that our collaborations continue through today and look forward to many more in the future.
\end{acknowledgement}
